---
title: "Session_1"
output: html_document
date: "2022-08-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
https://learningstatisticswithr.com/lsr-0.6.pdf
https://modernstatisticswithr.com/thebasics.html#runningcode
https://stat545.com/basic-data-care.html#buckle-your-seatbelt


# Basics of R

## Basic calculations

```{r}
4 + 5

62-54

4*5

62/4

2^5
```

## Variable assignment

```{r}
x <- 5

y <- "hello"

z <- c(1,2,3)

x
y
z

# calculations using variables

apples <- 5
oranges <- 8

apples + oranges
fruit <- apples + oranges
fruit

# shortcut for assignment operator: Alt -

```

Rules for names:

Cannot start with a digit or contain certain characters such as a comma or a space
Names are case-sensitive: data is a different variable to Data which is different to DATA

Conventions include using _ or .
data_full
data.full
dataFull

When we close R this will all be gone. For anything that we want to save and run again next time, we can write it in a script. To run code, press Ctrl + Enter on current line, or Ctrl + shift + Enter to run entire script.
Save it when done.

To make your code more readable, you can use comments using #. without this R tries to read it as code and you get an error, so need to use # so R will ignore it.
shortcut: ctrl + shift + c

# Vectors and data frames

We'll probably want to store more than one piece of information in one variable, for this we can use a vector.
```{r}
age <- c(29, 18, 34, 63, 56, 24, 53, 38)
```
Numbers are called elements.
We can then easily do something to each element in the vector:

```{r}
age.months <- age*12
age.months
```

Can access specific elements using their index
```{r}
age[2]
```
and change them
```{r}
age[2] <- 28
age
```
We can combine vectors into a table called a data frame.

```{r}
age.dataframe <- data.frame(age, age.months)
age.dataframe
```
# Functions in R

Once you have some data, you actually want to do something with it. To do so, you can use a function. This is just a ready-made set of instructions telling R what to do with the code.

```{r}
mean(age)
```
```{r}
round(3.14159)

round(3.14159, 2)
```
Here we have two arguments. In this function it's easy to remember which one means what, but in more complicated functions this isn't the case. That's why arguments can also have names, which allows us to specify arguments by name:

```{r}
round(x = 3.14159, digits = 2)
```
If you use argument names, it doesn't matter what order you type them in, but if you don't use names, you have to use the correct order. 
To find out what arguments a function takes:

```{r}
?round
```
Some arguments also have default values - e.g. in the first example, if you don't specify digits it will round to nearest whole number. So round has a default digits argument of 0.



In this instance, mean is the function and age is the argument. A function can have any number of arguments from none upwards.

# Packages

Packages are collections of functions and datasets that you can use in R but aren't automatically there at the beginning.

```{r}
install.packages("ggplot2")
library(ggplot2)
```
```{r}

```




# workspace and working directory

At some point you will want to leave R and maybe come back to it later. Or run multiple analyses that you want to keep separate. Or communicate your analyses in R to the outside world.
You need somewhere to store your work in R. 
can see where this currently is by typing getwd().
Recommended to do this via 'Projects'. 
File > New Project
When you close R it will ask if you want to save your workspace. Might seem tempting but generally not recommended as can cause problems further down the line. Can disable this in tools > global options.

# Importing Data

# Data cleaning




